---
title: "Intro to time series wrangling and viz"
author: "Vanessa Rathbone"
date: "2/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts)
library(slider)

```

## Read in the data

```{r}
toolik <- read_csv(here("data", "toolikweather.csv"))

#reading as characters
#quick plot to check it out if we don't change the characters
#it sees the dates as characters, not as the same variables
#dates as characters = NO BUENO

#ggplot(toolik, aes(x = date, y = mean_airtemp)) +
#geom_line()

```

### Convert this into a tsibble

```{r}

toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  as_tsibble(key = NULL, index = date)
```

```{r}

ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) +
  geom_line()

#trend = stationary, no major changes, syclicality = not really obvious if existing at all, seasonality is dominante
```

## Use index_by() function to aggregate time series information by intervals we specify

```{r}
#index_by() is the equivalent to group_by() but for time series data
#group by year, month, max dates, min dates, etc. 
#calculate the mean air temp 

toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% #"~yearmonth(.)" means function of yearmonth across all of the time series (.)
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))
  
ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~month(yr_mo, label = TRUE))

```

Aggregate by week
```{r}

toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE),
            min_airtemp = min(mean_airtemp))

head(toolik_weekly)
```

## Filtering time series

Use `filter_index()` to specify ranges of time series to keep or exclude.

```{r}

#filter by month/year
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")

#filter by day
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")

```

## Seasonplots

```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>% #or filter_index("2015" ~ "(.)")
  gg_season(y = mean_airtemp)

toolik_month %>% 
  gg_season(y = monthly_mean_temp) +
  theme_light()

#no notable 

```

## Seasonal subseries plot 
```{r}

toolik_month %>% 
  gg_subseries(monthly_mean_temp)
```

## Find moving average

Using the slider package in R

```{r}
set.seed(2021)
test <- rnorm(n = 100, mean = 40, sd = 10)
#made vector test

#look at each observation that the average will be calculated
slide(test, ~.x, .before = 2, .after = 2)

```

```{r}
#moving averages for 15 windows
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, 
                                   mean, 
                                   .before = 7,
                                   .after = 7)))

ggplot() +
  geom_line(data = toolik_ts, aes(x = date, y = mean_airtemp), 
            size = 0.2, 
            color = "grey") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d)) +
  theme_minimal()
```

## Create the ACF (autocorrelation function)

```{r}
#days time series

toolik_ts %>% 
  ACF(mean_airtemp) %>% 
  autoplot()

#lag in months on the x-axis, not a lot of dampening happening. 
toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()

```

## Decomposition by STl

```{r}
toolik_de <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

#not useful because all stored as an STL model
head(toolik_de)

components(toolik_de) %>% 
  autoplot()

#notice y-axis is is "free" so be sure to look at that. For example, trend looks like there's variability, but it's only between -6 to -10

```

